---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Fedorak, Alexandra"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: We have both qualitative and quantitative variables here, with Sex and Class being qualitative and the remaining variables all quantitative. It looks as though Sex is normally distributed with similar values across the 3 categories, but Class seems to have a bell-curve like distribution peaking at class A3. This could cause difficulties when testing amongst points on the tails of the distributions. Finally Volume has an incredibly large range with a minimum of 3.162 and a maximum of 995.673. This could make it difficult to compare volume to other variables.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
tab1 <- table(mydata$SEX, mydata$CLASS)
addmargins(tab1)

freq <- data.frame(tab1)
colnames(freq) <- c('Sex', 'Class', 'Frequency')
ggplot(freq, aes(x=Class, y=Frequency, fill=Sex, group=Sex)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  ggtitle("Abalone by Class")

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: It definitely looks as though sex distribution of abalones is not consistent throughout the classes. Classes A1 and A2 have a high majority of infants, but this seems to decrease significantly in classes A3-A5 where we see higher frequencies of Female and Male abalones. As for the distribution of the abalones by class, the class with the highest number of abalones is A3. From A1 to A5, there is somewhat a bell curve of abalones between the classes, peaking at A3 and having the lowest frequencies at A1 and A5.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
randomsample <- sample(seq(1:1036), 200)
work <- mydata[randomsample,]
plot(work[, 2:6])

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Whole weight (grams) vs Volume (cm^3)')) +
  ylab("Whole weight (grams)") +
  xlab("Volume (cm^3)")

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1) +
  ggtitle('Shuck weight (grams) vs. Whole weight (grams)') +
  geom_abline(intercept=0, slope=max(mydata$SHUCK/mydata$WHOLE)) +
  ylab("Shuck (grams)") +
  xlab("Whole (grams)") 

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: There seems to be a greater degree of variability between the points in the first plot (whole weight vs. volume) than the second plot (shuck weight vs. whole weight). Although there are definite correlations shown in both graphs, this suggests a slightly stronger correlation between the variables in the second graph. The location of the different age classes are similar between both graphs, with class A1 closest to the origin and points moving further from the origin as the age increases.***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
histogram <- ggplot(mydata, aes(RATIO, fill=SEX)) +
  facet_grid(. ~ SEX, space="free") +
  geom_histogram(bins=14, color="black", size=.2) + 
  ggtitle("Histograms of Ratio") +
  theme(legend.position="none", axis.title.x=element_blank()) +
  ylab("Frequency")
  
boxplot1 <- ggplot(mydata, aes(y=RATIO, fill=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_boxplot(color="black", size=.2, outlier.size=2, outlier.shape=1) + 
  ggtitle("Boxplots of Ratio") +
  theme(legend.position="none", axis.title=element_blank())

qqplot <- ggplot(mydata, aes(sample=RATIO, color=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_qq(cex=2, shape=1) + 
  ggtitle("QQ Plots of Ratio") +
  geom_qq_line(color="black", cex=.5) +
  theme(legend.position="none") +
  ylab("Sample Quantities") +
  xlab("Theoretical Quantities")

grid.arrange(histogram, boxplot1, qqplot, nrow=3)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: All distributions here are skewed to the right and they appear to be mostly normally distributed. The positive skew in all graphs is most likely due to the extreme outliers we see in the data.***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
outi <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
outf <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
outm <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(outi, outf, outm))

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: As to the results I've gathered in 3b, it looks as though the highest number of outliers comes from the infants category. Additionally the classes of the outliers seem to be between classes A1-A3 with only one of the outliers falling into the class A5. I believe this may be due to low survival rates of outliers and thus them not making it into the higher age classes.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
boxvol <- ggplot(mydata, aes(x=CLASS, y=VOLUME, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Boxplots of Volume by Class") +
  theme(legend.position="none")

boxwhole <- ggplot(mydata, aes(x=CLASS, y=WHOLE, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Boxplots of Whole weight by Class") +
  theme(legend.position="none")

scatvol <- ggplot(data = mydata, aes(x = RINGS, y = VOLUME)) + 
  geom_point(size = .5,) +
  ggtitle("Volume vs Rings") +
  theme(legend.position="none")

scatwhole <- ggplot(data = mydata, aes(x = RINGS, y = WHOLE)) + 
  geom_point(size = .5,) +
  ggtitle("Whole weight vs Rings") +
  theme(legend.position="none")

grid.arrange(boxvol, boxwhole, scatvol, scatwhole, nrow=2)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: I believe that neither of these variables would perform as good predictors of age. The boxplots showing volume by class and whole weight by class do show correlations between the two variables, but they are very rough with classes A3-A5 being very hardly distinguishable. Volume vs rings and whole weight vs rings also shows too much variability between data points to be a proper predictor of age.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
aggreg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))

mvolume <- matrix(aggreg$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

mshuck <- matrix(aggreg$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

mratio <- matrix(aggreg$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

kable(mvolume, caption="Volume")
kable(mshuck, caption="Shuck")
kable(mratio, caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
colnames(aggreg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')
ratiovclass <- ggplot(data = aggreg, aes(x = Class, y = Ratio, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Ratio by Class")

volvclass <- ggplot(data = aggreg, aes(x = Class, y = Volume, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Volume by Class")

shuckvclass <- ggplot(data = aggreg, aes(x = Class, y = Shuck, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Shuck weight by Class")

grid.arrange(ratiovclass, volvclass, shuckvclass, nrow=3)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: These plots show that ratio generally decreases with abalone age while mean volume and mean shuck weight generally increase with age. A question this could raise is why the males seem to have such low mean volumes and mean shuck weights - how are they even lower than the infant data points? And why are the females substantially higher in both the mean volume and mean shuck weight plots?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
infvol <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Infant Volume by Rings") +
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")

adultvol <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Adult Volume by Rings") +
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")
  
infwhole <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Infant Whole weight by Rings") +
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

adultwhole <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  ggtitle("Adult Whole weight by Rings") +
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

grid.arrange(infvol, adultvol, infwhole, adultwhole, nrow=2)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays show a steady increase in growth in about the first half of an abalone's life, then leveling out for the remainder of its lifespan. Sometimes after weight has reached its peak it even lowers slightly instead of leveling out. Additionally, there is a great amount of variation in the latter half of the infant volume by rings graph, depicted by the differences and similarities when looking at the Q1, median and Q3 marks within the boxplots. When comparing the infant versus the adult graphs, the general shape and pattern of the graphs look quite similar and there is surprisingly little variation between the volumes and weights when taking into account the difference in ages.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: I believe that based on the data, the original study could have failed by not taking infant abalones into account. With such large variations between volumes even in the adult abalones, this great range would have increased even further when not taking the age of the abalones into account. Additionally, counting the rings of an abalone has proven to be a very difficult task and has a lot of error associated with it. This could cause for inaccurate data leading to the failure of the first study.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: If I were presented with an overall histogram and summary statistics from a sample of some population, there are many important questions to ask before determining that it is representative of the whole population. Firstly, I would ask if the sample were random. I would also ask about the sample size compared to the size of the population. Additionally, I would ask when this sample was taken to determine if it was still recent enough to be an accurate representation. I would lastly ask about the variables within the sample and whether they are applicable to all groups questioned.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: When analyzing data derived from observational studies, there are outside factors to take into account that may have affected the collected data. For instance, in an observational study, how can we be sure that all relevant points were "observed"? Human error and different baises can come into play strongly when analyzing observational studies. Unfortunately, causality cannot be determined from this type of study. But, correlations can be witnessed, leading to further studies which can eventually determine causality.***
